Computing in the Field of Rationals

### 2012-02-26 rational.tex
copied out the rationals paper
lots of restructuring

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
%! Gx = \Gamma^{\times}
%! G = \Gamma
%! [] = \Box
%! |-->* = \mapsto^{*}
%! |-->> = \mapsto_{\ggg}
%! |--> = \mapsto
%! |- = \vdash
%! ==> = \Longrightarrow
%! <== = \Longleftarrow
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! -o = \multimap
%! ::= = &::=&
%! /= = \neq
%! forall = \forall
%! exists = \exists
%! empty = \epsilon
%! langRev = \Pi
%! langRevT = \Pi^{o}
%! langRevEE = \Pi^{\eta\epsilon}
%! theseus = Theseus
%! * = \times

@@

\title{Functional Pearl --- Programming with \\ 
  Negative, Fractional, Square Root, and Imaginary Types} 
\authorinfo{Roshan P. James}
           {Indiana University}
           {rpjames@indiana.edu}
\authorinfo{Amr Sabry}
           {Indiana University}
           {sabry@indiana.edu}
\maketitle

\begin{abstract}
  Every functional programmer knows about sum and product types, $a+b$ and
  $a*b$ respectively. Negative, fractional, square root, and imaginary types,
  $a-b$, $a/b$, $\sqrt{a}$, and $a+ib$ respectively, are much less known and
  their computational interpretation is unfamiliar and often complicated. We
  show that in a programming model in which information is preserved (such as
  the model introduced in our recent paper on \emph{Information Effects}),
  these types have particularly intuitive and natural computational
  interpretations. Intuitively, values of negative types are values that flow
  ``backwards'' to satisfy demands, values of fractional types are values
  that represent first-class ``structural'' constraints, and values of square
  root and imaginary types are values that can be related by arbitrary
  algebraic constraints. The combination of these negative, fractional,
  square root, and imaginary types enable greater flexibility in programming
  by breaking complicated invariants into local ones that can be
  independently satisfied by a subcomputation. Furthermore, they allow a
  programmer to express a wide range of programming idioms, including
  higher-order functions, delimited continuations, speculative computation,
  logic programming, and more.
\end{abstract}

@@

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Algebraic Types: Intuition}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Additive Duality in {{langRev}} }

%%%%%%%%%%%%%%%
\subsection{Syntax}

@@

%%%%%%%%%%%%%%%%%
\subsection{Rules for {{eta}} and {{eps}} }

@@

%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Diagrams}

Function

@@

Function application

@@

Function composition

@@

Delimited continuation

@@

Trace

@@

Double Negation

@@

@@

\subsection{To think about}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Multiplicative Duality in {{langRev}} }

>>>>

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
%! Gx = \Gamma^{\times}
%! G = \Gamma
%! [] = \Box
%! |-->* = \mapsto^{*}
%! |-->> = \mapsto_{\ggg}
%! |--> = \mapsto
%! |- = \vdash
%! ==> = \Longrightarrow
%! <== = \Longleftarrow
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! -o = \multimap
%! ::= = &::=&
%! /= = \neq
%! @@ = \mu
%! forall = \forall
%! exists = \exists
%! empty = \epsilon
%! langRev = \Pi
%! langRevT = \Pi^{o}
%! langRevEE = \Pi^{\eta\epsilon}
%! theseus = Theseus
%! * = \times

@@

\title{Computing in the Field of Rationals}
\authorinfo{Roshan P. James}
           {Indiana University}
           {rpjames@indiana.edu}
\authorinfo{Amr Sabry}
           {Indiana University}
           {sabry@indiana.edu}
\maketitle

\begin{abstract}
  Every functional programmer knows about sum and product types, $a+b$
  and $a*b$ respectively. Negative and fractional types, $a-b$ and
  $a/b$ respectively, are much less known and their computational
  interpretation is unfamiliar and often complicated. We show that in
  a programming model in which information is preserved (such as the
  model introduced in our recent paper on \emph{Information Effects}),
  these types have particularly intuitive and natural computational
  interpretations. Intuitively, values of negative types are values
  that flow ``backwards'' to satisfy demands, values of fractional
  types are values that represent first-class ``structural''
  constraints.
  %% , and values of square root and imaginary types are values that can
  %% be related by arbitrary algebraic constraints.
  The combination of these negative, fractional, square root, and
  imaginary types enable greater flexibility in programming by
  breaking complicated invariants into local ones that can be
  independently satisfied by a subcomputation. Furthermore, they allow
  a programmer to express a wide range of programming idioms,
  including higher-order functions, delimited continuations,
  speculative computation, logic programming, and more.
\end{abstract}

@@

In a technical sense, this paper extends the language of isomorphisms
{{langRev}}, with duality. Unlike Linear logic \cite{Girard87tcs} and
other systems which have one notion of duality over additive and
multiplicative componenents, {{langRevEE}} has two notions of duality
-- an additive duality over the monoid {{(0, +)}} and a multiplicative
duality over the monoid {{(1, *)}}. Each axis of duality also give us
a function space and hence {{langRevEE}} has an additive function
space correspoding to a notion of control or backtracking and a
multiplicative function space corresponding to a notion of unification
or constraint satisfaction.

@@

(NOTE: Lets use this section here to establish a long of ideas and
intuition. None of these connections have to be formal, but they can
be concrete.)

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{ {{langRevEE}} }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Additive Duality in {{langRev}} }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Multiplicative Duality in {{langRev}} }

\section{Semantics}
%%%%%%%%%%%%%%%
\subsection{Syntax}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Rules for {{eta}} and {{eps}} }

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Constructions in {{langRevEE}} }

These constructions are applicable to the additive and the
multiplicative monoid.

\paragraph*{Functions}

@@

\paragraph*{Function application}

@@

\paragraph*{Function composition}

@@

\paragraph*{Delimited continuation}

@@

\paragraph*{Trace }

@@

\paragraph*{Principium Contradictiones}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Constructions Specific to Negation}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Constructions Specific to Multiplication}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{To think about}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Advanced Programming Examples}

\subsection{SAT solver}

\subsection{GoI machine}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Field of Algebraic Numbers}

The field of algebraic numbers consist of the rationals extended with
solutions to polynomials. In a limited sense {{langRevEE}} with
recursive types already include polynomials such as the definition of
binary tree {{@@x.(1+x*x)}} implies the equality {{t = 1 + t*t}} and
hence {{t^2-t+1=0}}. While this equation does indeed have an
irrational imaginary root, as we will see -- {{langRevEE}} does not
have the means of expressing meaningful computation that uses the
root.

TODO.
